---
title: "SCI Calculation 3 ways"
author: "Emma Jones"
date: "5/18/2023"
output: html_document
---


This script is a minimal reprex to allow data external to the Virginia DEQ CEDS EDAS system to be run through the VDEQ SCI calculations (VSCI, VCPMI + 63, VCPMI - 65). This script will analyze benthic data provided it matches the required templates and the taxa are contained within the example master taxa list.
Please contact Emma Jones (emma.jones@deq.virginia.gov) if you have questions about these procedures.

NOTE (teaching repo adaptation): source/data paths use the repository structure (src/ and data/raw/)
 so the workflow runs after "Download ZIP" without moving files.
NOTE: Rarefaction to 110 individuals is applied to match the standard EDAS-style workflow
when sample totals exceed 110. Samples with totals ≤110 are unchanged.

## Setup
First, set up your environment.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
# built in R 3.6.2 but will work in newer R versions
library(tidyverse)
library(lubridate)
library(rmarkdown)
library(lazyeval)
```

## Important functions

These run the VSCI/VCPMI and are necessary to have in your environment. One day soon these will be distributed in a package. 

```{r}
# Teaching repo adaptation: explicit paths for ZIP-download reliability
source("src/VSCI_metrics_GENUS.R")
source("src/VCPMI_metrics_GENUS.R")
source("src/Updated_rarifyFunction.R")

# SCI wrapper function (runs one of the three index options)
SCI <- function(stationBenthicsDateRange, SCIchoice, benSamps, masterTaxaGenus, vmast){

  edas_options <- select(masterTaxaGenus, Class, Subclass, Order, Suborder, Superfamily, Family,
                         `Final VA Family ID`, FinalID) %>%
    mutate(across(where(is.factor), as.character))

  edas_list <- select(edas_options, `Final VA Family ID`, FinalID)

  # Lists for identifying taxonomy level in FinalID field
  GenusNames       <- unique(edas_options$FinalID)[!is.na(unique(edas_options$FinalID))]
  FamilyNames      <- unique(edas_options$Family)[!is.na(unique(edas_options$Family))]
  SuperfamilyNames <- unique(edas_options$Superfamily)[!is.na(unique(edas_options$Superfamily))]
  SuborderNames    <- unique(edas_options$Suborder)[!is.na(unique(edas_options$Suborder))]
  OrderNames       <- unique(edas_options$Order)[!is.na(unique(edas_options$Order))]
  SubclassNames    <- unique(edas_options$Subclass)[!is.na(unique(edas_options$Subclass))]
  ClassNames       <- unique(edas_options$Class)[!is.na(unique(edas_options$Class))]

  # Standardize benthic input
  EDASrare <- stationBenthicsDateRange %>%
    mutate(Count = Individuals) %>%
    select(BenSampID, FinalID, Count, `Excluded Taxa`) %>%
    mutate(
      GenusTaxaLevel       = ifelse(FinalID %in% GenusNames, T, F),
      FamilyTaxaLevel      = ifelse(FinalID %in% FamilyNames, T, F),
      SuperfamilyTaxaLevel = ifelse(FinalID %in% SuperfamilyNames, T, F),
      SuborderTaxaLevel    = ifelse(FinalID %in% SuborderNames, T, F),
      OrderTaxaLevel       = ifelse(FinalID %in% OrderNames, T, F),
      SubclassTaxaLevel    = ifelse(FinalID %in% SubclassNames, T, F),
      ClassTaxaLevel       = ifelse(FinalID %in% ClassNames, T, F)
    )

  # Backtrack FinalID to Final VA Family ID
  EDASrare2 <- left_join(EDASrare, edas_list, by = "FinalID") %>%
    filter(!is.na(`Final VA Family ID`)) %>%
    rename(`Genus Level Excluded Taxa` = `Excluded Taxa`)

# Rarefy to 110 individuals per sample (applies only when totals exceed 110)
# NOTE: rarify() returns columns: UID, finalID, SampleCount, Count
EDASrare2 <- rarify(
  bugdatlong = EDASrare2,
  UID        = "BenSampID",
  finalID    = "Final VA Family ID",
  count      = "Count"
) %>%
  rename(
    BenSampID = UID,
    `Final VA Family ID` = finalID
  )

  # Apply exclusion logic and aggregate to family level
  exclusionMath <- EDASrare2 %>%
    mutate(
      `Family Level Excluded Taxa` =
        ifelse(`Genus Level Excluded Taxa` == -1,
               ifelse(SuperfamilyTaxaLevel == TRUE | SuborderTaxaLevel == TRUE | OrderTaxaLevel == TRUE |
                        SubclassTaxaLevel == TRUE | ClassTaxaLevel == TRUE,
                      -1, 0),
               0)
    ) %>%
    group_by(BenSampID, `Final VA Family ID`) %>%
    summarise(
      `Family Level Count` = sum(Count),
      `Family Level Taxa` = n(),
      `Family Level Excluded Taxa` = sum(`Family Level Excluded Taxa`),
      `Final Family Level Taxa` = `Family Level Taxa` + sum(`Family Level Excluded Taxa`),
      .groups = "drop"
    )

  # Join traits/metric lookup
  bugTraits <- left_join(exclusionMath, vmast, by = c("Final VA Family ID"))

  # Run selected index option
  if (SCIchoice == "VSCI") {
    SCIout <- VSCIcalculation(bugTraits, exclusionMath, vmast) %>%
      mutate(SCI = "VSCI", `SCI Threshold` = 60) %>%
      rename("SCI Score" = "Fam SCI")
  }

  if (SCIchoice == "VCPMI + 63") {
    SCIout <- VCPMI63calculation(bugTraits, exclusionMath, vmast) %>%
      mutate(SCI = "VCPMI + 63", `SCI Threshold` = 42) %>%
      rename("SCI Score" = "CPMI63+CHOWAN")
  }

  if (SCIchoice == "VCPMI - 65") {
    SCIout <- VCPMI65calculation(bugTraits, exclusionMath, vmast) %>%
      mutate(SCI = "VCPMI - 65", `SCI Threshold` = 42) %>%
      rename("SCI Score" = "CPMI65-CHOWAN")
  }

  # Attach station metadata
  SCIout <- left_join(SCIout, benSamps, by = "BenSampID")
  return(SCIout)
}

```

## Master Taxa list

Bring in a master taxa list that contain genus level ID with a backtrack to family level as well as genus and 
family tolerance values, FFG, habit, etc. We will also manipulate it into a long format that will ensure we can 
run the SCI calculations efficiently. 

```{r}
masterTaxaGenus <- read.csv("data/raw/masterTaxaGenus.csv") %>%
  rename(`Final VA Family ID` = "Final.VA.Family.ID")

vmast <- masterTaxaGenus %>%
  rename(
    GenusTolVal = TolVal,
    TolVal      = FamTolVal,
    GenusFFG    = FFG,
    FFG         = FamFFG,
    GenusHabit  = Habit,
    Habit       = FamHabit
  ) %>%
  mutate(
    e  = ifelse(Order == "Ephemeroptera", 1, 0),
    p  = ifelse(Order == "Plecoptera",   1, 0),
    t  = ifelse(Order == "Trichoptera",  1, 0),
    tmin = ifelse(
      (Order == "Trichoptera" & Family != "Hydropsychidae") |
        (Order == "Trichoptera" & is.na(Family)),
      1, 0
    ),
    ept     = ifelse(e + p + t > 0, 1, 0),
    scraper = ifelse(FFG == "Scraper", 1, 0),
    chiro   = ifelse(Family == "Chironomidae", 1, 0),
    ptmin   = ifelse(p + tmin > 0, 1, 0),
    `clinger-HS` = ifelse(Habit == "Clinger" & !Family %in% c("Hydropsychidae", "Simuliidae"), 1, 0)
  ) %>%
  select(`Final VA Family ID`, TolVal, e, p, t, ept, ptmin, scraper, chiro, `clinger-HS`) %>%
  distinct(`Final VA Family ID`, .keep_all = TRUE) %>%
  filter(!is.na(`Final VA Family ID`)) %>%
  pivot_longer(-`Final VA Family ID`, names_to = "metric", values_to = "metric_val") %>%
  filter(!is.na(metric_val))

```
## Sample data 

Bring in benthic sampling information that matches this template.  This can contain multiple sites and sample events. 

```{r}
stationBenSamps <- readr::read_csv("data/raw/stationInfoBenSampsTESTSITE.csv")
```

## Benthic Data

Bring in benthic data as a separate object. This can contain multiple sites and sample events. 


```{r}
stationBenthics <- readr::read_csv("data/raw/stationBenthicsTESTSITE.csv")
```

## SCI

## Run one dataset through three index options (VSCI, VCPMI +63, VCPMI -65)

NOTE (teaching repo adaptation): we save each index output as a separate object and write
separate CSVs so students/labs can compare indices without overwriting results in memory.

```{r}
# Teaching repo adaptation: save each index output separately to avoid overwriting results.
dir.create("data/processed", showWarnings = FALSE, recursive = TRUE)

# --- 1) VSCI ---
SCIresults_vsci <- SCI(stationBenthics, "VSCI", stationBenSamps, masterTaxaGenus, vmast) %>%
  mutate_if(is.numeric, round, digits = 2) %>%
  arrange(`Collection Date`)
write.csv(SCIresults_vsci, "data/processed/VASCI_rarefied.csv", row.names = FALSE)

# --- 2) VCPMI + 63 ---
SCIresults_vcpmi63 <- SCI(stationBenthics, "VCPMI + 63", stationBenSamps, masterTaxaGenus, vmast) %>%
  mutate_if(is.numeric, round, digits = 2) %>%
  arrange(`Collection Date`)
write.csv(SCIresults_vcpmi63, "data/processed/VCPMI63_rarefied.csv", row.names = FALSE)

# --- 3) VCPMI - 65 ---
SCIresults_vcpmi65 <- SCI(stationBenthics, "VCPMI - 65", stationBenSamps, masterTaxaGenus, vmast) %>%
  mutate_if(is.numeric, round, digits = 2) %>%
  arrange(`Collection Date`)
write.csv(SCIresults_vcpmi65, "data/processed/VCPMI65_rarefied.csv", row.names = FALSE)

# --- Quick verification (helpful for students) ---
cat(
  "Wrote outputs to data/processed/:\n",
  "- VASCI_rarefied.csv\n",
  "- VCPMI63_rarefied.csv\n",
  "- VCPMI65_rarefied.csv\n"
)
```

      