---
title: "Instructor: SCI Calculation + Assignment Answer Key Starter"
author: "Sergio Sabat-Bonilla (teaching adaptation); original SCI code: Emma Jones (VDEQ)"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = TRUE
)

suppressPackageStartupMessages({
  library(tidyverse)
  library(lubridate)
  library(rmarkdown)
  library(lazyeval) # required by rarify() implementation
})

```

## Purpose (Instructor-only)

This instructor Rmd adds two robustness steps that make the workflow run cleanly across machines and data-entry conventions:

1) **Excluded Taxa recode:** Some CSVs store `Excluded Taxa` as `0/1`, while the original logic expects `0/-1`.  
2) **Preserve metadata across rarefaction:** `rarify()` returns only IDs + `SampleCount` + `Count`, so we **save and re-join** the exclusion flag and taxa-level flags needed downstream.

> Students can run the simpler “public” workflow. Instructors can use this file for grading, troubleshooting, and generating answer keys.

---

## Load required scripts

```{r}
source("src/VSCI_metrics_GENUS.R")
source("src/VCPMI_metrics_GENUS.R")
source("src/Updated_rarifyFunction.R")
```

Read inputs
```{r}
masterTaxaGenus <- read.csv("data/raw/masterTaxaGenus.csv") %>%
  rename(`Final VA Family ID` = "Final.VA.Family.ID")

stationBenSamps <- readr::read_csv("data/raw/stationInfoBenSampsTESTSITE.csv")
stationBenthics <- readr::read_csv("data/raw/stationBenthicsTESTSITE.csv")
```

Build vmast (traits/metrics lookup)
```{r}
vmast <- masterTaxaGenus %>%
  rename(
    GenusTolVal = TolVal,
    TolVal      = FamTolVal,
    GenusFFG    = FFG,
    FFG         = FamFFG,
    GenusHabit  = Habit,
    Habit       = FamHabit
  ) %>%
  mutate(
    e  = ifelse(Order == "Ephemeroptera", 1, 0),
    p  = ifelse(Order == "Plecoptera",   1, 0),
    t  = ifelse(Order == "Trichoptera",  1, 0),
    tmin = ifelse(
      (Order == "Trichoptera" & Family != "Hydropsychidae") |
        (Order == "Trichoptera" & is.na(Family)),
      1, 0
    ),
    ept     = ifelse(e + p + t > 0, 1, 0),
    scraper = ifelse(FFG == "Scraper", 1, 0),
    chiro   = ifelse(Family == "Chironomidae", 1, 0),
    ptmin   = ifelse(p + tmin > 0, 1, 0),
    `clinger-HS` = ifelse(Habit == "Clinger" & !Family %in% c("Hydropsychidae", "Simuliidae"), 1, 0)
  ) %>%
  select(`Final VA Family ID`, TolVal, e, p, t, ept, ptmin, scraper, chiro, `clinger-HS`) %>%
  distinct(`Final VA Family ID`, .keep_all = TRUE) %>%
  filter(!is.na(`Final VA Family ID`)) %>%
  pivot_longer(-`Final VA Family ID`, names_to = "metric", values_to = "metric_val") %>%
  filter(!is.na(metric_val))
```

SCI wrapper (robust instructor version)
```{r sci_instructor_function}
SCI_instructor <- function(stationBenthicsDateRange, SCIchoice, benSamps, masterTaxaGenus, vmast){

  edas_options <- masterTaxaGenus %>%
    dplyr::select(
      Class, Subclass, Order, Suborder, Superfamily, Family,
      `Final VA Family ID`, FinalID
    ) %>%
    mutate(across(where(is.factor), as.character))

  edas_list <- edas_options %>% dplyr::select(`Final VA Family ID`, FinalID)

  GenusNames       <- unique(edas_options$FinalID)[!is.na(edas_options$FinalID)]
  FamilyNames      <- unique(edas_options$Family)[!is.na(edas_options$Family)]
  SuperfamilyNames <- unique(edas_options$Superfamily)[!is.na(edas_options$Superfamily)]
  SuborderNames    <- unique(edas_options$Suborder)[!is.na(edas_options$Suborder)]
  OrderNames       <- unique(edas_options$Order)[!is.na(edas_options$Order)]
  SubclassNames    <- unique(edas_options$Subclass)[!is.na(edas_options$Subclass)]
  ClassNames       <- unique(edas_options$Class)[!is.na(edas_options$Class)]

  EDASrare <- stationBenthicsDateRange %>%
    mutate(Count = Individuals) %>%
    dplyr::select(BenSampID, FinalID, Count, `Excluded Taxa`) %>%
    mutate(
      GenusTaxaLevel       = FinalID %in% GenusNames,
      FamilyTaxaLevel      = FinalID %in% FamilyNames,
      SuperfamilyTaxaLevel = FinalID %in% SuperfamilyNames,
      SuborderTaxaLevel    = FinalID %in% SuborderNames,
      OrderTaxaLevel       = FinalID %in% OrderNames,
      SubclassTaxaLevel    = FinalID %in% SubclassNames,
      ClassTaxaLevel       = FinalID %in% ClassNames
    )

  EDASrare2 <- left_join(EDASrare, edas_list, by = "FinalID") %>%
    filter(!is.na(`Final VA Family ID`)) %>%
    mutate(`Genus Level Excluded Taxa` = ifelse(`Excluded Taxa` == 1, -1, 0))

  keep_cols <- EDASrare2 %>%
    dplyr::select(
      BenSampID, `Final VA Family ID`, `Genus Level Excluded Taxa`,
      GenusTaxaLevel, FamilyTaxaLevel, SuperfamilyTaxaLevel,
      SuborderTaxaLevel, OrderTaxaLevel, SubclassTaxaLevel, ClassTaxaLevel
    ) %>%
    group_by(BenSampID, `Final VA Family ID`) %>%
    summarise(across(everything(), ~ max(as.numeric(.x), na.rm = TRUE)), .groups = "drop") %>%
    mutate(across(
      c(GenusTaxaLevel, FamilyTaxaLevel, SuperfamilyTaxaLevel,
        SuborderTaxaLevel, OrderTaxaLevel, SubclassTaxaLevel, ClassTaxaLevel),
      ~ as.logical(.x)
    ))

  EDASrare2_r <- rarify(
    bugdatlong = EDASrare2,
    UID        = "BenSampID",
    finalID    = "Final VA Family ID",
    count      = "Count"
  )

  EDASrare2_r <- left_join(EDASrare2_r, keep_cols, by = c("BenSampID", "Final VA Family ID"))

  exclusionMath <- EDASrare2_r %>%
    mutate(
      `Family Level Excluded Taxa` =
        ifelse(`Genus Level Excluded Taxa` == -1,
               ifelse(SuperfamilyTaxaLevel | SuborderTaxaLevel | OrderTaxaLevel |
                        SubclassTaxaLevel | ClassTaxaLevel,
                      -1, 0),
               0)
    ) %>%
    group_by(BenSampID, `Final VA Family ID`) %>%
    summarise(
      `Family Level Count` = sum(Count),
      `Family Level Taxa` = n(),
      `Family Level Excluded Taxa` = sum(`Family Level Excluded Taxa`),
      `Final Family Level Taxa` = `Family Level Taxa` + sum(`Family Level Excluded Taxa`),
      .groups = "drop"
    )

  bugTraits <- left_join(
    exclusionMath, vmast,
    by = c("Final VA Family ID"),
    relationship = "many-to-many"
  )

  if (SCIchoice == "VSCI") {
    SCIout <- VSCIcalculation(bugTraits, exclusionMath, vmast) %>%
      mutate(SCI = "VSCI", `SCI Threshold` = 60) %>%
      rename(`SCI Score` = `Fam SCI`)
  } else if (SCIchoice == "VCPMI + 63") {
    SCIout <- VCPMI63calculation(bugTraits, exclusionMath, vmast) %>%
      mutate(SCI = "VCPMI + 63", `SCI Threshold` = 42) %>%
      rename(`SCI Score` = `CPMI63+CHOWAN`)
  } else if (SCIchoice == "VCPMI - 65") {
    SCIout <- VCPMI65calculation(bugTraits, exclusionMath, vmast) %>%
      mutate(SCI = "VCPMI - 65", `SCI Threshold` = 42) %>%
      rename(`SCI Score` = `CPMI65-CHOWAN`)
  } else {
    stop("SCIchoice must be one of: 'VSCI', 'VCPMI + 63', 'VCPMI - 65'")
  }

  left_join(SCIout, benSamps, by = "BenSampID")
}
```

Run all three indices + write outputs
```{r}
dir.create("data/processed", showWarnings = FALSE, recursive = TRUE)

SCI_vsci    <- SCI_instructor(stationBenthics, "VSCI",       stationBenSamps, masterTaxaGenus, vmast)
SCI_vcpmi63 <- SCI_instructor(stationBenthics, "VCPMI + 63", stationBenSamps, masterTaxaGenus, vmast)
SCI_vcpmi65 <- SCI_instructor(stationBenthics, "VCPMI - 65", stationBenSamps, masterTaxaGenus, vmast)

write_csv(SCI_vsci,    "data/processed/VASCI_rarefied.csv")
write_csv(SCI_vcpmi63, "data/processed/VCPMI63_rarefied.csv")
write_csv(SCI_vcpmi65, "data/processed/VCPMI65_rarefied.csv")

cat("Wrote outputs to data/processed/\n")
```

Starter analyses to answer the assignment (Instructor key scaffolding)

Q1–Q2: Site summaries + threshold exceedance
```{r}
all_scores <- bind_rows(
  SCI_vsci    %>% select(StationID, BenSampID, `Collection Date`, SCI, `SCI Score`, `SCI Threshold`),
  SCI_vcpmi63 %>% select(StationID, BenSampID, `Collection Date`, SCI, `SCI Score`, `SCI Threshold`),
  SCI_vcpmi65 %>% select(StationID, BenSampID, `Collection Date`, SCI, `SCI Score`, `SCI Threshold`)
)

site_summary <- all_scores %>%
  group_by(StationID, SCI, `SCI Threshold`) %>%
  summarise(
    n = n(),
    mean = mean(`SCI Score`, na.rm = TRUE),
    sd   = sd(`SCI Score`, na.rm = TRUE),
    var  = var(`SCI Score`, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(above_threshold = mean >= `SCI Threshold`) %>%
  arrange(SCI, StationID)

site_summary
```

Q3: Time series (VSCI example)
```{r vsci_object}
vsci <- SCI_vsci %>%
  mutate(`Collection Date` = as.Date(`Collection Date`))
```

Plot A — “true time series” (stations with repeated samples)
```{r}
vsci_rep <- vsci %>%
  count(StationID) %>%
  filter(n >= 3) %>%    # set to 2 or 3 depending on how strict you want
  pull(StationID) %>%
  (\(ids) filter(vsci, StationID %in% ids))()

ggplot(vsci_rep, aes(x = `Collection Date`, y = `SCI Score`)) +
  geom_line(linewidth = 0.9, alpha = 0.85) +
  geom_point(size = 2, alpha = 0.9) +
  geom_hline(yintercept = 60, linetype = "dashed") +
  facet_wrap(~ StationID, scales = "free_x") +
  labs(
    title = "VSCI over time (stations with repeated sampling only)",
    x = "Collection date",
    y = "VSCI score"
  ) +
  theme_bw()
```

Plot B — Single-visit  and seasonal sites grouped by season instead
```{r q3b_single_visit_by_season}
# Sites with only one visit (no time series possible)
single_ids <- vsci %>%
  count(StationID) %>%
  filter(n == 1) %>%
  pull(StationID)

vsci_single <- vsci %>%
  filter(StationID %in% single_ids) %>%
  mutate(
    Month = month(`Collection Date`),
    Season = case_when(
      Month %in% c(12, 1, 2)  ~ "Winter",
      Month %in% c(3, 4, 5)   ~ "Spring",
      Month %in% c(6, 7, 8)   ~ "Summer",
      Month %in% c(9, 10, 11) ~ "Fall"
    ),
    Season = factor(Season, levels = c("Spring","Summer","Fall","Winter"))
  )

ggplot(vsci_single, aes(x = Season, y = `SCI Score`)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.25) +
  geom_jitter(width = 0.15, alpha = 0.8, size = 2) +
  geom_hline(yintercept = 60, linetype = "dashed") +
  labs(
    title = "VSCI single-visit sites by season",
    x = "Season (from collection date)",
    y = "VSCI score"
  ) +
  theme_bw()
```

Plot C — CRO seasonal sites combined (CRO1 Fall + Spring in one panel)
This turns CRO1Fall and CRO1Spring into one base site CRO1, with Season as the grouping.
```{r q3c_cro_fall_spring_combined}
vsci_cro <- vsci %>%
  filter(str_detect(StationID, "^CRO")) %>%
  mutate(
    Season = case_when(
      str_detect(StationID, "Fall$")   ~ "Fall",
      str_detect(StationID, "Spring$") ~ "Spring",
      TRUE ~ NA_character_
    ),
    SiteBase = str_replace(StationID, "(Fall|Spring)$", ""),
    Season = factor(Season, levels = c("Spring","Fall"))
  ) %>%
  filter(!is.na(Season))

ggplot(vsci_cro, aes(x = Season, y = `SCI Score`, group = Season)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.25) +
  geom_jitter(width = 0.12, alpha = 0.85, size = 2) +
  geom_hline(yintercept = 60, linetype = "dashed") +
  facet_wrap(~ SiteBase) +
  labs(
    title = "VSCI CRO sites: Spring vs Fall (combined by base site)",
    x = "Season (from StationID suffix)",
    y = "VSCI score"
  ) +
  theme_bw()
```

Q8: Total abundance vs VSCI (example)
```{r}
abund <- stationBenthics %>%
  group_by(BenSampID) %>%
  summarise(TotalIndividuals_raw = sum(Individuals, na.rm = TRUE), .groups = "drop")

q8_df <- SCI_vsci %>%
  left_join(abund, by = "BenSampID")

ggplot(q8_df, aes(x = `SCI Score`, y = TotalIndividuals_raw)) +
  geom_point(alpha = 0.85) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Total abundance (raw) vs VSCI",
    x = "VSCI score",
    y = "Total individuals (raw)"
  ) +
  theme_bw()
```

Q9: Rarefaction effect (optional if you keep a “pre-rarefaction” branch)

If you want a true before/after, you’ll need to compute VSCI once without rarefaction and compare.
(This instructor file currently runs the rarefied workflow to match EDAS.)